---
title: "Initial QC & Filtering"
author: "Friederike Dündar"
date: "6/27/2019; updated `r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_depth: 6
        code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, cache.lazy = FALSE, warning = FALSE, eval=FALSE)
```

```{r cache=FALSE, warning = FALSE, message = FALSE, eval=TRUE}
library(data.table)
library(magrittr)
library(ggplot2); theme_set(theme_bw(base_size = 16))

condition_cols <- c("cyan4","blue2","gray85")
cell_pop_cols <- c("goldenrod4","gold1")
```

There was a total of three sequencing rounds to obtain the desired depth.

> Fastq files for all 3 runs were aggregated and aligned to GRCz10.

Since they sequenced multiple times, the sequencing depth is fairly high, i.e. the UMI cut-offs that
CellRanger typically uses may not work as well. In fact, for the cells of the "all" set, 
there seems to be a large portion of cell-free droplets that's assigned to be a cell
(CellRanger estimates are 15K+).

Originally, there were 6 samples:

* whole heart: Control, 3 days and 7 days after injury
* *tcf21*-sorted (epicardial cell enrichment): Control, 3 days and 7 days after injury

## Initial QC

```{r reading_in_10X_results, eval=FALSE}
# conda activate muR --> conda environment with latest R installation
library(scater)
library(magrittr)
library(Matrix)
library(scABC2) # min.  v. 1.5.0. to provide function for reading in 10X data

samples <- c("3dpa-all","3dpa-epi","7dpa-all","7dpa-epi","Ctrl-all","Ctrl-epi")

data_dirs <- paste0("2019_06_Jingli/data/CellRanger/", samples, "/outs/filtered_feature_bc_matrix/")
names(data_dirs) <- samples
sce <- read10XResults.abc(data_dir=data_dirs,
                           barcode_file="barcodes.tsv.gz", genes_file = "features.tsv.gz", matrix_file = "matrix.mtx.gz",
                           expand = TRUE, min_total_cell_counts = 0, min_mean_gene_counts=0) ##capture all the cells/genes that CR deemed worth keeping

########################## Initial Filtering ###############################
dim(sce)
#[1] 32520 57744
keep_gns <- rowSums(counts(sce)) >0
table(keep_gns)
# keep_gns
# FALSE  TRUE 
#  6626 25894 
sce <- sce[keep_gns,]
dim(sce)
#[1] 25894 57744

## add chromosome information -----------------------------------
library(biomaRt)
ensembl = useMart("ensembl",dataset="drerio_gene_ensembl")
chrs <- getBM(attributes=c("ensembl_gene_id", "chromosome_name"), mart = ensembl) 
row.names(chrs) <- chrs$ensembl_gene_id
chrs <- chrs[rownames(sce),]

rowData(sce)$chr <- paste0("chr", chrs[rownames(sce),]$chromosome_name)
sce <- scater::calculateQCMetrics(sce,  feature_controls = list(mito = which(chrs$chromosome_name == "MT")))

colData(sce)$condition <- gsub("-.*", "", sce$Sample)
colData(sce)$cell_populations <- gsub(".*-", "", sce$Sample)

##! save(sce, file = "2019_06_Jingli/data/sce_original.rda")


########################## PART 2 CELL FILTERING ###############################
library(ABCutilities) # v. 0.3.0
library(ggplot2); theme_set(theme_bw(base_size = 16))
condition_cols <- c("cyan4","blue2","gray85")
cell_pop_cols <- c("goldenrod4","gold1")

source("2019_06_Jingli/src/qc_plots.R")

## Mito content ---------------------------------------------------------
high.mito <- lapply(samples, function(x) isOutlier(sce[,sce$Sample == x]$pct_counts_mito, nmads = 5, type = "higher"))
names(high.mito) <- samples
lapply(samples, function(x) max(sce[,sce$Sample == x ][, !high.mito[[x]]]$pct_counts_mito)) %>% unlist
#[1] 52.459016  8.198198 67.151899  6.472081 71.195652  5.575540

## Number of genes per cell--------------------------------------------
outlier.genes <- lapply(samples, function(x) isOutlier(sce[, sce$Sample == x]$total_features_by_counts, nmads = 3, type = "both"))
names(outlier.genes) <- samples
lapply(samples, function(x) summary( sce[, sce$Sample == x][, !outlier.genes[[x]]]$log10_total_features_by_counts))

# [[1]]
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   1.820   2.516   2.723   2.741   2.955   3.296 
# 
# [[2]]
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   1.491   2.997   3.216   3.169   3.438   3.726 
# 
# [[3]]
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   1.908   2.565   2.765   2.782   3.025   3.302 
# 
# [[4]]
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   1.431   2.964   3.087   3.060   3.203   3.446 
# 
# [[5]]
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   1.568   2.479   2.651   2.632   2.846   3.160 
# 
# [[6]]
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   1.322   2.920   3.032   2.998   3.117   3.341 


outlier.umi <- lapply(samples, function(x) isOutlier(sce[, sce$Sample == x]$total_counts, nmads = 3, type = "both"))
names(outlier.umi) <- samples
lapply(samples, function(x) summary(sce[, sce$Sample == x][, !outlier.umi[[x]]]$log10_total_counts))
# [[1]]
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   2.717   3.039   3.286   3.315   3.583   4.008 
# 
# [[2]]
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   2.701   3.557   3.891   3.833   4.196   4.529 
# 
# [[3]]
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   2.829   3.173   3.425   3.443   3.712   4.065 
# 
# [[4]]
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   2.702   3.491   3.696   3.664   3.892   4.206 
# 
# [[5]]
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   2.764   3.157   3.328   3.343   3.548   3.862 
# 
# [[6]]
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   2.700   3.418   3.609   3.573   3.766   4.069

## making plots ----------------------
qc_params <- c("log10_total_features_by_counts", "log10_total_counts", "log10_total_counts_mito","pct_counts_mito")

#pdf("qc01_countsPerCell.pdf", width = 12)
for(i in samples){
  lapply(qc_params, function(x){
    print(x)
  
   thresh.list <- list( log10_total_features_by_counts = c(min(sce[, sce$Sample == i][,!outlier.genes[[i]]]$log10_total_features_by_counts),
                                                           max(sce[, sce$Sample == i][,!outlier.genes[[i]]]$log10_total_features_by_counts)), ## lower and upper threshold
                        log10_total_counts = c(min(sce[, sce$Sample == i][, !outlier.umi[[i]]]$log10_total_counts),
                                               max(sce[, sce$Sample == i][, !outlier.umi[[i]]]$log10_total_counts)),
                        log10_total_counts_mito = c(0,0), # dummy entries
                        pct_counts_mito = c(0, max(sce[, sce$Sample == i][, !high.mito[[i]]]$pct_counts_mito))
                     )

  
    P <- plot_qc_cells_violins(as.data.frame(colData(sce[, sce$Sample == i])), 
                             x_val = "Sample",
                             value_col = x,
                             col_val = "condition")#,
                             #threshold = thresh.list[[x]])
    #P <- P+ scale_color_manual(values = condition_cols)
    P <- P + guides(color = FALSE) + scale_color_manual(values = "black")
    png(paste0("qc01_",i, "_", x, ".png"), height = 200)
    print(P)
    dev.off()
  })
}
#dev.off()
```

Based on the initial QC results, it is apparent that the whole heart samples suffered from extensive cell death and contaminated droplets. 
**We therefore focused on the *tcf21*-sorted samples.**

### Cell filtering

Min. number of genes: `r 10^2.5` and max % mitochondrial reads: 5.

```{r}
thresh.list <- list( min_log10_total_features_by_counts = 2.5, 
                     max_pct_counts_mito = 5)
```

```{r}
pdf("qc01_countsPerCell.pdf", width = 12)

tl <- list( log10_total_features_by_counts = c(2.5, 3.5), ## lower and upper threshold
                        pct_counts_mito = c(0, 5))

  lapply(names(tl), function(x){
    print(x)
    P <- plot_qc_cells_violins(as.data.frame(colData(sce[, sce$cell_populations == "epi" ])), 
                             x_val = "condition",
                             value_col = x,
                             col_val = "condition",
                             threshold = tl[[x]])
    P <- P+ scale_color_manual(values = condition_cols)
    #P <- P + guides(color = FALSE) + scale_color_manual(values = "black")
    print(P)
  })
dev.off()
```


```{r}
## focus on epicardial samples
sce <- sce[, sce$cell_populations == "epi" ]
dim(sce)
#[1] 25894 16676

## filter cells ---------------------------------------------------------
qc_df <- as.data.frame(colData(sce))
filtered <- list()

keep_cells <- row.names(subset(qc_df, log10_total_features_by_counts >= thresh.list$min_log10_total_features_by_counts & pct_counts_mito <= thresh.list$max_pct_counts_mito))

## keep track of those that will be removed
filtered$cells01 <- row.names(subset(qc_df, log10_total_features_by_counts < thresh.list$min_log10_total_features_by_counts & pct_counts_mito > thresh.list$max_pct_counts_mito))

if(length(keep_cells) > 0){
  sce <- sce[,keep_cells]
}
dim(sce)
#[1] 25894 15141

## immediately remove no-coverage-genes --------------------------------------
gnsfilt1 <- scABC2::filter_gene_countBased(sce.object = sce, min.rowSums = 0, return_gns= TRUE)
filtered$genes_zero_after_cells01 <- gnsfilt1$remove_genes

if(length(gnsfilt1$keep_genes) >0){
  sce <- sce[gnsfilt1$keep_genes, ]
}
dim(sce)
```

### Gene Filtering

```{r eval=FALSE}
## make more sensible rownames ---------------------------
rownames(sce) <- uniquifyFeatureNames(rowData(sce)$id, rowData(sce)$symbol)

## update QC measures
sce <- scater::calculateQCMetrics(sce,  feature_controls = list(mito = which(rowData(sce)$chr == "chrMT")))

png("qc02_highestExpressedGenes.png", width = 600, height=700)
plotHighestExprs(sce) 
dev.off()


hk_genes <- unique(c(grep("^mt-", rownames(sce), value=TRUE, ignore.case=TRUE),  # mitochondrial genes
                     grep("^Rp[sl]", rownames(sce), value=TRUE, ignore.case=TRUE))) # ribosomal genes

png("qc02_highestExpressedGenes_noHKGenes.png", width = 600, height=700)
plotHighestExprs( sce[!(rownames(sce) %in% hk_genes), ] ) 
dev.off()

## calculate drop out rates ======================
gns_dropouts <- scABC2::calc_dropouts_per_cellGroup(sce, rownames(sce), split_by = "Sample")
##!save(gns_dropouts, file = "2019_06_Jingli/data/gns_dropouts.rda")

gns_dropouts$condition <- gsub("-.*","", gns_dropouts$Sample)

P <- ggplot(data = gns_dropouts,
       aes(x = log10(mean.pct.of.counts),
           y = log10(pct.zeroCov_cells + .1),
           text = paste(gene, condition, sep = "_"))) + 
  geom_point(aes(color = condition), shape = 1, size = .5, alpha = .5) +
  geom_point(data = gns_dropouts[gene %in% hk_genes],
             aes(fill = condition), shape = 22, size = 4, alpha = .8) +
  facet_grid(~condition) + ggtitle("Gene dropout rates", subtitle = "Squares = housekeeping genes") +
  facet_wrap(~Sample) +
  scale_color_manual(values = condition_cols) +
  scale_fill_manual(values = condition_cols)

png("qc02_gns_dropoutsPostCellFilt.png", width=1200, height = 900)
print(P)
dev.off()


cutoff <- -2.5
print(paste("Gene cutoff:", cutoff))

P <- ggplot(data = gns_dropouts,
       aes(x = log10(mean.pct.of.counts),
           y = log10(pct.zeroCov_cells + .1),
           text = paste(gene, condition, sep = "_"))) + 
  geom_point(aes(color = condition), shape = 1, size = .5, alpha = .5) +
  geom_point(data = gns_dropouts[gene %in% hk_genes],
             aes(fill = condition), shape = 22, size = 4, alpha = .8) +
  facet_grid(~condition) + ggtitle("Gene dropout rates -- zoomed in") +
  coord_cartesian(ylim = c(1.9,2.01), xlim = c(-3,-1)) +
  facet_wrap(~Sample) +
  geom_vline(xintercept = cutoff, linetype = "dashed", color = "grey") +
  scale_color_manual(values = condition_cols) +
  scale_fill_manual(values = condition_cols)

png("qc02_gns_dropoutsPostCellFilt_zoom.png", width=800, height = 800)
print(P)
dev.off()

keep_genes  <- gns_dropouts[ (Sample == "3dpa-epi" & mean.pct.of.counts >= 10^-2.75) | (Sample != "3dpa-epi" & mean.pct.of.counts >= 10^-2.5)]$gene %>% unique

sce.filt <- sce[keep_genes,]
dim(sce.filt)
#[1] 24127 15141

sce.filt <- scater::calculateQCMetrics(sce.filt, feature_controls = list(mito = which(rowData(sce.filt)$chr == "chrMT")))

qc_params_gns <- c("log10_mean_counts", "log10_total_counts", "pct_dropout_by_counts", "n_cells_by_counts")

for(i in unique(sce.filt$Sample)){
  sc.tmp <- sce.filt[, sce.filt$Sample == i]
  sc.tmp <- scater::calculateQCMetrics(sc.tmp, feature_controls = list(mito = which(rowData(sc.tmp)$chr == "chrMT")))
  qc_gns <- as.data.frame(rowData(sc.tmp))
  
  pl.gns <- lapply(qc_params_gns, function(x){
    plot_qc_gns_violins(qc_gns, value_col = x,
                        x_val = "is_feature_control") +
                        guides(color = FALSE) + xlab("control feature") +
      ggtitle(i)
    })
  
  png(paste0("qc03_gns_distributions_postFilt_",i, ".png"), width=600, height = 800)
  ABCutilities::MultiPlotList(pl.gns, title = paste("Gene assessments", i))
  dev.off()
  
  rm(sc.tmp);  rm(qc_gns);  rm(pl.gns);  gc()
}
```




# Session info

```
R version 3.5.1 (2018-07-02)
Platform: x86_64-conda_cos6-linux-gnu (64-bit)
Running under: Red Hat Enterprise Linux Server release 6.3 (Santiago)

Matrix products: default
BLAS/LAPACK: /miniconda3/envs/muR/lib/R/lib/libRblas.so

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
 [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
 [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
 [9] LC_ADDRESS=C               LC_TELEPHONE=C            
[11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

attached base packages:
[1] parallel  stats4    stats     graphics  grDevices utils     datasets 
[8] methods   base     

other attached packages:
 [1] Seurat_3.1.0                Matrix_1.2-17              
 [3] magrittr_1.5                scater_1.10.1              
 [5] ggplot2_3.2.1               SingleCellExperiment_1.4.1 
 [7] SummarizedExperiment_1.12.0 DelayedArray_0.8.0         
 [9] BiocParallel_1.16.6         matrixStats_0.54.0         
[11] Biobase_2.42.0              GenomicRanges_1.34.0       
[13] GenomeInfoDb_1.18.2         IRanges_2.16.0             
[15] S4Vectors_0.20.1            BiocGenerics_0.28.0        

loaded via a namespace (and not attached):
 [1] tsne_0.1-3               nlme_3.1-141             bitops_1.0-6            
 [4] httr_1.4.1               RcppAnnoy_0.0.12         RColorBrewer_1.1-2      
 [7] sctransform_0.2.0        tools_3.5.1              R6_2.4.0                
[10] irlba_2.3.3              KernSmooth_2.23-15       HDF5Array_1.10.1        
[13] vipor_0.4.5              uwot_0.1.3               lazyeval_0.2.2          
[16] colorspace_1.4-1         npsurv_0.4-0             withr_2.1.2             
[19] tidyselect_0.2.5         gridExtra_2.3            compiler_3.5.1          
[22] plotly_4.9.0             caTools_1.17.1.2         scales_1.0.0            
[25] lmtest_0.9-37            pbapply_1.4-2            ggridges_0.5.1          
[28] stringr_1.4.0            digest_0.6.19            R.utils_2.9.0           
[31] XVector_0.22.0           htmltools_0.3.6          pkgconfig_2.0.2         
[34] bibtex_0.4.2             htmlwidgets_1.3          rlang_0.4.0             
[37] DelayedMatrixStats_1.4.0 zoo_1.8-6                jsonlite_1.6            
[40] ica_1.0-2                gtools_3.8.1             dplyr_0.8.3             
[43] R.oo_1.22.0              RCurl_1.95-4.12          GenomeInfoDbData_1.2.0  
[46] Rcpp_1.0.1               ggbeeswarm_0.6.0         munsell_0.5.0           
[49] Rhdf5lib_1.6.0           ape_5.3                  reticulate_1.13         
[52] viridis_0.5.1            R.methodsS3_1.7.1        stringi_1.4.3           
[55] gbRd_0.4-11              MASS_7.3-51.4            zlibbioc_1.28.0         
[58] rhdf5_2.28.0             gplots_3.0.1.1           Rtsne_0.15              
[61] plyr_1.8.4               grid_3.5.1               gdata_2.18.0            
[64] listenv_0.7.0            ggrepel_0.8.1            crayon_1.3.4            
[67] lattice_0.20-38          cowplot_1.0.0            splines_3.5.1           
[70] SDMTools_1.1-221.1       pillar_1.4.2             igraph_1.2.4.1          
[73] future.apply_1.3.0       reshape2_1.4.3           codetools_0.2-16        
[76] leiden_0.3.1             glue_1.3.1               lsei_1.2-0              
[79] metap_1.1                RcppParallel_4.4.3       data.table_1.12.2       
[82] png_0.1-7                Rdpack_0.11-0            tidyr_0.8.3             
[85] gtable_0.3.0             RANN_2.6.1               purrr_0.3.2             
[88] future_1.14.0            assertthat_0.2.1         rsvd_1.0.2              
[91] survival_2.44-1.1        viridisLite_0.3.0        tibble_2.1.3            
[94] beeswarm_0.2.3           cluster_2.1.0            globals_0.12.4          
[97] fitdistrplus_1.0-14      ROCR_1.0-7  
```
